Twenty years of dynamic occupancy models: a review of applications and look towards the future

### 2025-03-04 Manuscript.qmd
Major changes to MS over the past two weeks prior to resubmission

abstract: | 
 Occupancy models.

format:
  html:
    theme: simplex
  docx: default

@@

Ecologists are often asked to describe, explain, and predict where and when species occur on the landscape. Research on patterns of species occupancy dates back to the origins of the field, and continues to a priority in quantitative ecological research [@humboldt1849]. Estimates of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions [@mackenzie2013]. Robust knowledge of the occupancy patterns of a species can also help us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios [@kéry2013].

While occupancy is a useful concept, it is also a challenging quantity to describe, measure, and estimate. The need to understand and quantitatively describe species occupancy has led to the development of several popular modelling approaches, including stochastic patch occupancy models commonly applied to study meta-population dynamics [@gutiérrez-arellano2024], and species distribution models (SDMs) widely used to explore species occurrence at larger scales [@franklin2010]. However, several factors can make occupancy difficult to estimate. For instance, simple presence/absence observations can be biased when species are detected imperfectly – this is often the case in wildlife field data, where it can be impossible to determine from a single observation whether a location is truly occupied or whether the species occurs but was not detected [@gu2004; @lahoz-monfort2014]. Despite the ubiquity of imperfect detection in data collection, many models fit to presence/absence data make no adjustment for this source of bias [@kellner2014]. Another challenge for modelling occupancy is the difficulty in describing populations under non-equilibrium conditions, where a species’ occurrence pattern and relationship to its environment is in flux [@dormann2007; @elith2010]. These conditions often occur during biological invasions and climate change driven range shifts, each of which are conservation priorities and increasingly common scenarios in the Anthropocene [@bertelsmeier2013; @lenoir2015].

The site occupancy models first introduced by @mackenzie2002 set the foundation for a powerful framework for modelling presence/absence data while accounting for each of these challenges [@guillera-arroita2017a] . Drawing on principles from the mark-recapture literature, occupancy models use multiple resurveys of sites to estimate detection probabilities and correct for bias in estimates of site occupancy. Originally a static model, @mackenzie2003 extended this model for use in multiple time-steps by explicitly describing the process of changing occupancy via colonisation and extinction, relaxing assumptions of equilibrium and allowing description of patterns of site occupancy through time. For newcomers to DOMs, in [Box @nte-Primer] we present a simple introduction to the basic model form as described in @mackenzie2003.

This dynamic occupancy model (DOM) balances complexity and feasibility, explicitly describing the key processes driving occupancy dynamics while requiring reasonably simple-to-collect presence absence/data instead of the detailed demographic or abundance data required by more process-explicit models [@briscoe2019]. These features make the DOM an important tool, with uses including hypothesis testing of relationships between occupancy and the environment, explorations of the key drivers of occupancy, and even prediction of occupancy under future conditions [@briscoe2021; @kéry2013].

@@

Dynamic occupancy models.

![](images/clipboard-1253177906.png){width="11cm"}

![](images/clipboard-3209412869.png){width="11cm"}
:::

DOMs make several key assumptions requiring careful consideration by model users, which we outline here as our review interrogates related aspects of model building.

I.  **False positive detections do not occur**. While this assumption can be safely met in many studies, it is not necessarily guaranteed when working with more cryptic species or less reliable survey methods. @mcclintock2010 and @miller2015 comment on the bias induced when false positives occur and are not accounted for, highlighting the need to consider how certain detections truly are for any given study system. Significantly, even genuine detections of a species can be considered 'false positives' when they do not represent the intended definition of occupancy, such as detections of transient individuals when the intent is to estimate breeding occupancy [@berigan2019]. Where this assumption can not reasonably be met, model extensions designed to account for false positive error should be considered [@royle2006; @miller2011; @miller2015].

II. **Sites are closed to occupancy between seasons.** This requirement, best known as the 'closure assumption,' has also been subject to considerable discussion in terms of the bias introduced when it is violated [@rota2009; @otto2013]. Closure is dependent not only on the life history of the species, but also on the definition of occupancy used by researchers — short seasons may represent dynamics more representative of species 'use,' and it can be difficult to distinguish local extinction from temporary emigration [@valente2017]. Model extensions to relax the closure assumption have been developed, including approaches using staggered arrival and departure periods between sites [@kendall2013]. A more pertinent approach for most studies, however, is careful consideration of the appropriate definition of occupancy under the survey design used [@steenweg2018].

III. **Heterogeneity in occupancy and detection is accounted for.** As with any approach for modelling species occurrence, it is assumed that DOMs appropriately capture variation in occupancy patterns and species detectability across the study system. Generally, this is achieved by allowing model parameters (ψ~1~, γ, ε, and ρ) to vary with respect to covariates representing the environmental factors which may be expected to influence these parameters. An important element of this assumption is that the likelihood of detection of a species can depend not only on the observability of the species, but also on factors like habitat suitability which influence species abundance and activity [@guillera-arroita2017a]. While no model will fully account for the complexity inherent in patterns of species occupancy and detection, failure to capture key drivers is likely to introduce bias and confound inference made from model estimates [@barry2006]. Compared to the first two assumptions mentioned, this aspect of DOMs has been less thoroughly discussed and comparatively little is known about how this latent heterogeneity can influence model performance.

Since its original description in @mackenzie2003, the DOM has been further developed with numerous model extensions and alternative formulations including aforementioned implementations accounting for false positives [@royle2006; @miller2011; @miller2015], multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimated multi-species models [@dorazio2010]. For a comprehensive discussion of common extensions and their applications see @bailey2014 and @guillera-arroita2017a, as well as @devarajan2020 for a more detailed review of multi-species occupancy models.

# Review methods

Methods.

@@

Coverage.

![Bars indicate the 92 articles included in our review as a proportion of the estimated number of published articles fitting DOMs, based on the qualification rate for articles in each strata. The proportion of articles included from each strata were as follows: 12% from 2005-2008; 24% from 2008-2011; 42% from 2012-2015; 35% from 2016-2019; and 57% from 2020-2023.](Figures/CoveragePlot.jpeg){#fig-coverage}

Stamp collection.

![A) Locations of study areas where data was collected for reviewed DOMs. B) Spatial extent of study areas, defined as the area of inference, within which all surveyed points were contained. C) Number of articles which fit models to each category of taxa. Taxa were considered 'threatened' if they are listed on the IUCN Red List, or if authors indicate that they are otherwise threatened. D) Explicitly multi-species models include both hierarchical, jointly estimated models as well as more interactive models. Some studies fit both independent and multi-species models, such that these values do not sum to our sample size. E) Survey methods used to collected presence/absence data. Note that some articles employed multiple detection methods, and that some methods (e.g., citizen bird counts) may fall into multiple categories. F) Quantity of sites where surveys were conducted and duration of studies. Yellow bars indicate median values for site quantity (100) and study duration (8.2 years). Study duration is defined as the time elapsed between the first and last survey.](Figures/ReviewBox.png){#fig-StudyDetails}

Covariates.

::: column-page-inset
```{r covariate table}
#| echo: false
#| message: false
#| label: tbl-covariates
#| tbl-cap: "All covariates considered for inclusion in a study were classified into mutually exclusive categories. We calculate the percentage of studies which include at least one covariate of a given category on any parameter, Initial Occupancy (&psi;1), Occupancy (&psi;), Colonisation (&gamma;), Extinction(&epsilon;), and Detection (&rho;). We also present the average percentage of covariates in a category which are dynamic (varying through seasons) and directly observed, as well as the percentage of articles which model each category of covariate with a non-linear relationship or as part of an interaction with another covariate."

library(tidyverse)
library(gt)

covTable <- read_csv("Figures/CovariateTable.csv")

covTable |>
  mutate(Summary = case_when(Covariate %in% c("Environmental", "Structural", "Overall") ~ TRUE,
                             .default = FALSE)) |>
  arrange(Category, Summary, -Prop_Any) |>
  gt(rowname_col = "Covariate", groupname_col = "Category") |>
  tab_options(
    row_group.as_column = TRUE
  ) |>
  tab_row_group(
    label = md("Environmental covariates"),
    rows = Category == "E",
    id = "Env"
  ) |>
  tab_row_group(
    label = md("Structural covariates"),
    rows = Category == "S",
    id = "Struct"
  ) |>
  tab_row_group(
    label = md("All covariates"),
    rows = Category == "O",
    id = "All"
  ) |>
  tab_style(
    style = cell_fill(color = "aquamarine4"),
    locations = cells_row_groups(groups = "Env")
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "plum4")
    ),
    locations = cells_row_groups(groups = "Struct")
  ) |>
  tab_style(
    style = list(
      cell_fill(color = "gray60")
    ),
    locations = cells_row_groups(groups = "All")
  ) |>
  row_group_order(groups = c("Env", "Struct", "All")) |>
  tab_spanner(label = "Percentage of articles with covariate on parameters:",
              columns = c("Prop_Any", "Initial occupancy", "Occupancy",
                          "Colonisation", "Extinction_Persistence",
                          "Detection"),
              id = "inclusions") |>
  tab_spanner(label = "Percentage which are:",
              columns = c(Prop_Dynamic, Prop_Direct),
              id = "covariates") |>
  tab_spanner(label = "Articles representing this covariate with:",
              columns = c(Prop_Nonlinear, Prop_Interact),
              id = "representations") |>
   cols_label(
    Prop_Any = "Any",
    `Initial occupancy` = html("&psi;1"),
    Occupancy = html("&psi;"),
    Colonisation = html("&gamma;"),
    Extinction_Persistence = html("&epsilon;"),
    Detection = html("&rho;"),
    Prop_Dynamic = "Dynamic",
    Prop_Direct = "Directly observed",
    Prop_Nonlinear = "Non-linear relationships",
    Prop_Interact = "Interactions between covariates"
  ) |>
  text_case_match(
    "Environmental" ~ html("<b>Any Environmental</b>"),
    "HABT" ~ html("<b>Habitat </b>\n<i>Aspects of habitat and land cover</i>"),
    "PHEN" ~ html("<b>Phenology</b>\n<i>Time-varying elements distinct from sampling occasions</i>"),
    "SPAT" ~ html("<b>Spatial</b>\n<i>Site geometry, connectivity, or other spatial elements</i>"),
    "OTHER-E" ~ html("<b>Other environmental</b>\n<i>Other environmental covariate not otherwise listed</i>"),
    "TOPO" ~ html("<b>Topography</b>\n<i>Elements of landscape topography</i>"),
    "HYDR" ~ html("<b>Hydrology</b>\n<i>Aspects of hydrology, such as distance to water</i>"),
    "CLWE" ~ html("<b>Climate/Weather</b>\n<i>Climate, weather, and natural disasters</i>"),
    "ANTH" ~ html("<b>Anthropogenic</b>\n<i>Relations to human activity</i>"),
    "BIOT" ~ html("<b>Biotic interaction</b>\n<i>Interactions with other (non-plant) species</i>"),
    "Structural" ~ html("<b>Any Structural</b>"),
    "SEAS" ~ html("<b>Primary occasion</b>\n<i>Effect of the primary occasion</i>"),
    "OBSE" ~ html("<b>Observation</b>\n<i>Details on the observation process</i>"),
    "SURV" ~ html("<b>Secondary occasion</b>\n<i>Effect of the secondary occasion</i>"),
    "SPEC" ~ html("<b>Species effect</b>\n<i>Species-level effects</i>"),
    "SITE" ~ html("<b>Site effect</b>\n<i>Site-level effects</i>"),
    "OTHER-S" ~ html("<b>Other structural</b>\n<i>Other structural covariate not otherwise listed</i>"),
    "Overall" ~ html("<b>All covariates</b>"),
    .locations = cells_stub()
  ) |>
  cols_hide(c(Category, Summary)) |>
  fmt_percent(decimals = 0) |>
  sub_missing(missing_text = "-") |>
  data_color(
    columns = where(is.numeric),
    method = "numeric",
    palette = "viridis",
    alpha = 0.5
  ) |>
  tab_style(locations = cells_stub(rows = "Environmental"),
            style = cell_fill(color = "aquamarine4", alpha = 0.5)) |>
  tab_style(locations = cells_body(rows = "Environmental"),
            style = list(cell_fill(color = "aquamarine4", alpha = 0.5),
                         cell_text(weight = "bold", style = "italic",
                                   color = "black"))) |>
  tab_style(locations = cells_stub(rows = "Structural"),
            style = cell_fill(color = "plum4", alpha = 0.5)) |>
  tab_style(locations = cells_body(rows = "Structural"),
            style = list(cell_fill(color = "plum4", alpha = 0.5),
                         cell_text(weight = "bold", style = "italic",
                                   color = "black"))) |>
  tab_style(locations = cells_stub(rows = "Overall"),
            style = cell_fill(color = "gray60", alpha = 0.5)) |>
  tab_style(locations = cells_body(rows = "Overall"),
            style = list(cell_fill(color = "gray60", alpha = 0.5),
                         cell_text(weight = "bold",
                                   color = "black"))) |>
  tab_style(locations = cells_row_groups(),
            style = cell_text(color = "white")) |>
  as_raw_html()
```
:::

More covariates.

![The number of covariates considered for each parameter across all studies in our sample. 'Occupancy' given here represents the alternative parameterisation of the DOM which jointly estimates Occupancy for every season, Colonisation, and Detection, where Extinction is a derived parameter; this differs from the more popular Initial occupancy/Colonisation/Extinction/Detection parameterisation. Here, a 'covariate' is defined as each distinct covariate considered for inclusion. Linear and quadratic representations of the same covariate are counted as one covariate.](Figures/CovParamPlot.jpeg){#fig-covariates}

Modelling summary.

```{r Modelling summary table}
#| echo: false
#| message: false
#| label: tbl-modelling
#| tbl-cap: "Modelling practices in dynamic occupancy models, subset by frequentist or bayesian implementations. The median covariate count presented here represents the median quantity of covariates considered for each model parameter across the studies in our review. The model selection methods represented in this table are non-exclusive and some articles employ multiple approaches. 2 models included in the 'Overall' column are neural network based and fall into neither the Frequentist or Bayesian categories."

>>>>

abstract: | 
 Since their introduction over twenty years ago, dynamic occupancy models (DOMs) have become a powerful and flexible framework for estimating species occupancy across space and time while accounting for imperfect detection. As their popularity has increased and extensions have further expanded their capabilities, DOMs have been applied to increasingly diverse datasets and research objectives in applied ecology. At the same time, technological advancements have resulted in massive increases in available data, offering both new opportunities and new challenges for users of DOMs. Given these developments, it is timely to examine common practices in building these models to understand the breadth of modelling approaches, determine potential vulnerabilities, and identify priorities for future research. We reviewed a sample of articles that have fit DOMs in the past 20 years, examining the contexts of their application and the approaches taken to the model building process. We find that these models have been used to pursue diverse objectives, based on datasets with wide-ranging spatial and temporal scales collected using a variety of survey methods. Our comparisons of modelling approaches indicate that many applications of DOMs considered relatively few covariates on key model parameters, as well as a tendency towards linear responses over more complex non-linear or interactive forms. Model selection techniques were largely idiosyncratic with little consensus on the best approaches, and model evaluation was rare across reviewed applications – additionally, existing research on either of these subjects is limited. Based on these findings, we identify key areas of the modelling process which merit discussion and further investigation. Where possible we provide recommendations for current users of DOMs, and where uncertainties remain, we highlight key priorities for future research to support users in fitting the most reliable and useful models possible.

format:
  html:
    theme: simplex
    mainfont: Georgia, serif
  docx: default

@@

The description of patterns of species occupancy across landscapes has been a long-standing subject of ecological research [@humboldt1849]. Estimates of how widespread a species is and where it occurs are the foundation of monitoring programs and important for assessing conservation status, while identifying potential drivers of occurrence can help inform potential management actions [@mackenzie2013]. Robust knowledge of the occupancy patterns of a species can also help us to predict where a species is most likely to occur, both under present conditions and in hypothetical future scenarios [@kéry2013].

Irrespective of its importance, ‘occupancy’ – broadly referring here to the presence of a species in a given area and time period – has proven to be a persistently difficult quantity to estimate in practice. This difficulty arises from the challenges inherent in modelling complex and dynamic natural systems with incomplete data, exacerbated by the fact that it is often impossible to determine whether a species is truly absent from a site or whether it was simply not detected. It is well-established that imperfect detection of organisms results in biased estimates of occupancy [@gu2004; @lahoz-monfort2014]. Despite this fact and the ubiquity of imperfect detection in field data, many models in widespread use make no adjustments for detection [@kellner2014]. In addition to this limitation, many models currently in use (such as conventional correlative species distribution models) are ill-suited for predicting to new locations and time periods [@dormann2007; @elith2010]. These shortcomings are particularly pronounced when modelling biological invasions and climate change driven range shifts, both of which are management priorities that are increasingly commonplace in the Anthropocene [@bertelsmeier2013; @lenoir2015].

In recent years, awareness of these obstacles to accurate occupancy estimation and the limitations of many popular methods has encouraged interest in a variety of process-explicit models better suited to these contexts [@briscoe2019]. These include ‘dynamic occupancy models’ (DOMs), first introduced by @mackenzie2003 as an extension to earlier static site occupancy models [@mackenzie2002]. In their simplest form DOMs use hierarchical observation data to estimate changing site occupancy over time while correcting for imperfect detection; an overview of the basic form and requirements of these models can be found in Box 1. By accounting for imperfect detection and explicitly modelling colonisation and extinction to describe changes in site occupancy DOMs provide answers to common issues in occupancy estimation and offer important advantages over other models [@guillera-arroita2017a]. Alongside these advantages, they require only relatively simple-to-collect detection/non-detection data rather than the detailed demographic or abundance data required by other more process-explicit models [@briscoe2019]. This balance between complexity and feasibility makes DOMs an appealing option for many applied ecologists.

@@

DOMsare hierarchical models that link observed detection/non-detection data with the underlying latent process of changing site occupancy. To do this, they contain two [sub-models]{.underline}: one describing the ecological process of sites shifting between unoccupied and occupied states over time, and one describing the observation process that records whether a species is detected during surveys given its presence at a site.

**Input data:** To separate these two processes DOMs require hierarchical data inputs. Occupancy is estimated at independent sites *(i)* during discrete, time-bound intervals termed ‘primary occasions’ *(t)*. Within primary occasions*,* multiple observations are made at the same sites during independent ‘secondary occasions’ *(j).* For each observation, a species is either detected (y~i,t,j~ = 1) or not (y~i,t,j~ = 0), resulting in a three dimensional detection matrix as shown in the example below. It is not necessary for all primary occasions to contain the same number of secondary occasions, and missing observations at sites can be accommodated. While secondary occasions are most often represented as repeat observations, other options including same-visit replicates and spatial replicates may also be used [@guillera-arroita2017a].

![](Figures/B1_RobustDesign.png){fig-align="center"}

**Occupancy sub-model:** Changing site occupancy over time is described as a Markovian process as represented in the graphic below. In each primary occasion *t,* each site *i* may exist in either an occupied (z~i,t~ = 1) or unoccupied state (z~i,t~ = 0). In the first primary occasion, occupancy is determined by a Bernoulli trial with initial occupancy probability 𝛙~1~, such that $z_{i,t=1} \sim Bernoulli(\psi_1)$. In subsequent primary occasions, site occupancy is determined by the site's state in the previous primary occasion and probabilities of colonisation 𝛄 and extinction 𝛆, such that $z_{i,t} \sim Bernoulli(z_{i,t-1}(1-\epsilon)+(1-z_{i,t-1})*\gamma)$.

![](Figures/B2_Occupancy.png){fig-align="center"}

**Observation sub-model:** The detection process relates the observed data **y** to the latent occupancy states **z** as shown in the below graphic. During any given secondary occasion *j* within primary occasion *t* at site *i,* observers will either observe (y~i,t,j~ = 1) or not observe (y~i,t,j~ = 0) the target species. The probability of observation is given as a Bernoulli trial with detection probability *p,* conditioned on the site being occupied in primary occasion *t* such that $y_{i,t,j} \sim Bernoulli(z_{i,t}*p)$.

![](Figures/B3_Detection.png){fig-align="center"}

In their original formulation, DOMs estimate four parameters of interest: initial occupancy (𝛙~1~), colonisation (𝛄), extinction (𝛆), and detection (*p)*. While we have represented each of these parameters as constants for simplicity, they are more often modelled as time-dependent or estimated with respect to covariates. These covariates are typically incorporated via a logit link function, such that DOMs can be considered four interconnected generalised linear models.

Conventional DOMs make several important assumptions: i) sites are closed to changes in true occupancy state within each primary occasion, ii) occupancy at each site is independent of other sites, iii) observations within each primary occasion are independent of each other, iv) no false positive detections occur, and v) key sources of heterogeneity (including in detectability) are modelled.

For more detailed information on DOMs and related models, see [@mackenzie2017] and [@kéry2021].
:::

Over the past two decades, continued research on DOMs and occupancy models more broadly has established a powerful and flexible modelling framework suitable for many common tasks in ecological research. Uptake of these models has been encouraged by the availability of freely available software tools for fitting DOMs, which include the program PRESENCE and the R package *unmarked* for fitting the models by maximum likelihood estimation [@hines2006; @kellner2023]*.* In the Bayesian context, resources such as @kéry2021 ’s text on hierarchical modelling in JAGS and the *ubms* R package have helped to increase the accessibility of these implementations [@kellner2022]. Various model extensions have further broadened DOMs’ capabilities – these include models that can account for false positives [@royle2006; @miller2011], model multiple states beyond occupied and unoccupied [@nichols2007], and jointly estimate occupancy for multiple species [@dorazio2010; @devarajan2020]. For more details on using these extensions, see reviews by @bailey2014 and @guillera-arroita2017a.

Coinciding with these developments, recent years have seen substantial changes in how ecologists conduct their research. The amount of data available for modelling (including species detections as well as environmental data) has grown considerably over time due to a range of factors, including improvements to data sharing, new large-scale monitoring programs, and increased interest in citizen science efforts [@farley2018; @altwegg2019]. Technological advances have facilitated the widespread deployment of autonomous detection methods including camera traps and acoustic monitors, generating large quantities of observation data suitable for analysis with DOMs [@balantic2019; @lahoz-monfort2021]. At the same time, advancements in computing have made methods which may have been too computationally expensive in the past far more accessible for many. While these advances create exciting new opportunities, they also introduce new challenges for users who must navigate a complex model building process to produce useful and reliable models. Where researchers on related species distribution models have built a large body of literature on assessing various approaches to model building including covariate inclusion and model selection, comparatively little research on these topics has been published for DOMs despite additional complications in performing these tasks for hierarchical models [@guisan2017].

### Objectives

To understand how users have applied DOMs over the past two decades, we conducted an in-depth review of studies implementing these models since their introduction. We first identify the research contexts where DOMs have been used, characterising the research objectives they have been applied to, the scale and characteristics of the study systems where data was collected, and the methods used to collect and format the data used to fit models. We then review approaches taken to the modelling process, including the nature of the covariates considered for inclusion and the form of their relationships in the model, the methods used for model selection, and the reporting of model assessment and evaluation. By jointly considering the research contexts to which DOMs are now applied and the approaches taken for model building and evaluation, we aim to highlight challenges in building DOMs, providing recommendations where possible and identifying priorities for future research where uncertainties remain.

# Methods

Our review was constrained to applications of the dynamic occupancy model of @mackenzie2003 and its extensions. To be included, articles must have fit models that met the following criteria:

i.   Used non-simulated, field-collected, detection/non-detection data.

ii. Included data from multiple sites which could exist in at least two states, including an occupied and unoccupied state.

iii. Included data from multiple primary occasions, between which sites change states conditional on the prior occasion’s occupancy state and transition probabilities such as colonisation and extinction.

iv. Contained at least one parameter describing the detection process.

A pool of candidate articles was generated using two queries on Web of Science on July 26 and July 29, 2024. The first query included all articles from 2004-2023 which cited @mackenzie2003. To capture additional relevant articles that did not directly cite @mackenzie2003, a second query was generated to search articles in the same time-span matching the terms “*dynamic occupancy model\**”, “*multi-season occupancy model\**”, or “*occupancy dynamic\**”; articles including each of “*occupancy*”, “*colonization*”, “*extinction/persistence*”, and “*detection*”; and articles with the term “*occupancy*” located near “*dynamic*” in the title, key terms, or abstract. These queries resulted in 1469 articles: 897 retrieved only from the MacKenzie citations, 274 only from the keywords search, and 298 which appeared in both queries. To allow comparison of DOMs through time, we divided all articles across five four-year strata spanning 2004-2007, 2008-2011, 2012-2015, 2016-2019, and 2020-2023. From each stratum we randomly selected 20 articles for inclusion in the review. Articles that did not meet inclusion criteria were replaced from within their own stratum unless no articles remained. For each article we documented details on the research contexts, datasets, and modelling processes as outlined below.

### Research objectives

We allocated each article to research objective categories based on the aims stated in the text. The six non-exclusive categories included i) ‘estimating trends’, where authors expressed interest in estimates of site occupancy, colonisation, extinction, or detection probabilities; ii) ‘testing relationships’, where authors examined predefined hypothesised relationships between covariates and model parameters; iii) ‘identifying drivers’, where authors more broadly explored covariates associated with model parameters; iv) ‘predicting temporally’, where authors predicted site occupancy under future conditions; v) ‘predicting spatially’, where authors predicted site occupancy to unsurveyed locations; and vi) ‘developing methods’, where authors introduced, tested, or demonstrated aspects of DOMs.

### Study systems

We recorded the approximate geographical location, spatial scale, and target taxa for each study system in the reviewed articles. In cases where a single article included DOMs fit to data from multiple study systems, we recorded details for each dataset that was analysed independently. A study system’s spatial scale was defined as the intended area of inference containing all sites, measured to an order of magnitude to account for uncertainty in reporting. Details collected on target taxa included the total number of taxa modelled, their general categorisation (birds, mammals, herptiles, invertebrates, or ‘other’), and their conservation status. Taxa were denoted as threatened if they were listed on the IUCN Red List of Species as of 2024 or if authors explicitly stated that they were threatened [@iucn]. This deference to authors’ representation of conservation status was made to account for sub-species which lack listings or species which are of more local concern, although we acknowledge that in some cases conservation status may have changed since publication.

### Observations and data structure

For each article, we recorded how observation data was collected for use in modelling. Categories of survey methods included human observations, physical trapping, camera traps, and acoustic monitors. Within these categories we also indicated whether any observations were collected by citizen scientists, either as part of structured survey programs or as more *ad hoc* observations. Details collected on each dataset’s structure included the number of primary and secondary occasions, the time elapsed between the first and last survey, and the number of sites used for modelling.

### Covariates and complexity

We were interested in the types and quantities of the covariates considered by authors. To this end, we recorded all covariates considered in each study regardless of whether they were included in final models, acknowledging that not all covariates considered are always reported. Key traits of each covariate were recorded including a general categorisation (see supporting information), whether they were directly observed or remotely sensed, whether they were static or varied between primary occasions, and how they were represented in the model: as a linear response, a non-linear response, or part of an interaction with another covariate [@james2021]. In some cases, a single article included distinct modelling workflows with different candidate covariates. Covariates were recorded for each of these workflows independently.

### Model selection and assessment

Model selection procedures were sorted into non-exclusive categories including ‘*a priori’*, where only one model was considered; ‘candidate set’, where a predefined set of models was considered; ‘sequential’, where covariates were selected parameter-by-parameter (e.g., fitting detection first, followed by initial occupancy and so on); and ‘simple precursors’, where covariate selection was preceded by tests with a simpler model implementation such as a linear regression or single season occupancy model. We also indicated whether model averaging was used for multi-model inference @burnham2004. For each modelling workflow, we documented whether goodness-of-fit was tested and reported, and whether model performance was assessed by validation with either in-sample or out-of-sample data.

@@

92 articles were included for this review, fewer than the 100 possible articles due to a deficit of qualifying papers in the first stratum (2004-2007). All review data is available in the supporting information. Based on the acceptance rates within each stratum and the number of unprocessed articles remaining for each year, an estimated 496 of 1152 unprocessed articles in our sample would have met inclusion criteria, with an apparent increase over time in the number of articles fitting DOMs (@fig-coverage).

![Bars indicate the estimated number of articles fitting DOMs in each year, with green representing articles included in our review and grey projected remaining qualifying articles. Projections were calculated using the number of unprocessed articles remaining in each year multiplied by the acceptance rate of the corresponding stratum: 12% of queried articles from 2004-2007; 24% from 2008-2011; 42% from 2012-2015; 35% from 2016-2019; and 57% from 2020-2023.](Figures/1_CoveragePlot.png){#fig-coverage}

### Research objectives

DOMs have been used to achieve varied research objectives, with no one category of objective representing over half of usage and 37% of studies having pursued multiple objectives simultaneously (Figure 2). The most frequent use of DOMs (44% of studies) was to test hypothesised relationships between environmental factors and species occupancy, often targeting core conservation priorities for their focal species as demonstrated by @olson2005 ’s early study assessing the influence of barred owls (*Strix varia)* on the threatened spotted owl (*Strix occidentalis*). 35% of studies took a more exploratory approach to identifying possible drivers of occupancy, such as in @huber2017 ’s study that examined dozens of habitat covariates for wood warbler (*Phylloscopus sibilatrix*) occupancy. An additional 36% of articles used DOMs to monitor trends in occupancy state through time, both for single species of high conservation interest and for broader community assemblages of species [@ahumada2013; @scott2015]. While only 17% of articles used their DOMs to make predictions, the proportion of articles pursing this objective has increased in recent years. These studies tended to have a strong conservation focus, as with @mcgowan2020 ’s projections of future occupancy for wetland birds under alternative management scenarios. Finally, 22% of the papers reviewed focused on methods development, representing continued focus on extending and testing DOMs.

The applied nature of studies fitting DOMs is further reflected in their authorship: while 79% of studies included at least one affiliate of an academic institution, 59% included a government affiliate, 28% an NGO affiliate, and 12% an affiliate of a private company. 63% of studies had cross-sector authorship with affiliations from at least two categories.

![Bars represent the percentage of articles which pursued each of six non-exclusive research objectives. These objectives include articles which demonstrated methods for DOMs, estimated trends in model parameters, identified drivers of occupancy dynamics, tested hypothesised relationships with the environment, made spatial predictions, and made temporal predictions. Percentages are given for each four-year stratum (n = 12 articles for 2004-2007; n = 20 for all other strata), as well as for all articles in the review sample (n = 92).](Figures/2_Objectives.png){#fig-Objectives}

### Study systems

The study systems included in our sample were globally distributed: while a majority of articles used data collected in the United States of America, study systems spanned all continents except Antarctica (@fig-StudyDetails A). These study areas varied considerably in size, ranging from hyper-local to continental scales (@fig-StudyDetails B). The smallest study locality included in our sample explored insect occurrence in a rainforest plot less than one square kilometre, while the largest analysis modelled avian range shifts across the entire eastern half of the United States [@clement2019; @basset2023].

Most studies focused on vertebrate taxa (@fig-StudyDetails C), particularly birds and mammals. DOMs have been less frequently applied to non-animal organisms, perhaps due to a reduced emphasis on imperfect detection outside of the wildlife modelling community. However, there were exceptions, including studies that used DOMs to model decadal changes in lichen occupancy, the spread of chronic wasting disease, and mosquito dynamics [@belinchón2017; @mores2020; @cook2022]. The vast majority of studies model terrestrial species, though there have been a limited number of studies on aquatic taxa including invasive salmon, Great Plains stream fishes, and whales [@falke2012; @fisher2014; @pendleton2022].

While most studies fit models for a single taxon (either a single species or multiple species lumped together), 44% fit models to multiple distinct taxa (@fig-StudyDetails D). 34% of studies fit independent models to multiple species [@otto2012; @peach2019], whereas 12% used explicitly multi-species extensions of DOMs. The latter included hierarchical models which fit hundreds of species in a single implementation with species-level effects [@dorazio2010; @hendershot2020] as well as explicit interaction models which estimated conditional occupancy, colonisation, extinction, and detection probabilities [@lesmeister2015; @fidino2019].undefined

![A summary of the research contexts in reviewed applications of DOMs. a) The approximate locations of the study areas where data was collected. b) Spatial extent of the study areas, defined as the area of inference within which all surveyed points were contained. c) Number of articles that fit models to each taxa category. Taxa are considered ‘threatened’ if indicated by authors or they are listed on the IUCN Red List. d) The number of taxa modelled as distinct groups in each application. Explicitly multi-species models include both hierarchical jointly estimated models and more interactive models. e) Number of articles using each method for capturing detection/non-detection data. These are non-exclusive, and citizen science data is a subset of ‘direct observations.’ (f) The number of sites and study duration for each dataset used to fit DOMs – gold lines indicate medians.](Figures/3_StudyContexts.png){#fig-StudyDetails}

### Observations and data collection

The detection/non-detection data required for DOMs was collected using various survey methods across reviewed studies (@fig-StudyDetails E): 79% of studies used direct observation data, 11% used live-trapping methods, and 14% used detections from camera traps. Within these categories, 10% of studies used citizen science data. Citizen science data included coordinated surveys at backyard bird-feeders as well as interviews with civilians on sightings of tiger (*Panthera tigris*) signs [@zuckerberg2011; @warrier2020]. Datasets ranged considerably in size, with the smallest including observations at just 10 sites and the largest including over 6000 (@fig-StudyDetails F); the median dataset included 100 sites. The temporal scale of studies shows similar variability – time elapsed between the first and last survey ranged from under one month to over forty years (median 8 years), and the number of primary occasions ranged from 2 to 189 (median 6 occasions).

Notably, not all of these datasets were originally collected in a hierarchical structure with DOMs in mind. In these cases, authors formatted their data into a hierarchical format *post-hoc* using a variety of methods. Some defined primary occasions as arbitrary discrete time intervals, treating all surveys occurring within a window as secondary occasions; others defined sites as larger grid cells, treating any survey falling within the grid as a spatial replicate. For example, in the only application of DOMs to marine species in our sample, @pendleton2022 used aerial transects broken up into grid cells to observe whale occupancy. In another example with grid cells, @marescot2020 fit a multi-species model treating poachers as a taxon and using ranger reports from each cell to create detection histories. These manipulations permit use of data predating DOMs, with one study using surveys conducted by Joseph Grinnell in 1908 to model century-long changes in occupancy [@riddell2021].

### Covariates and complexity

The most common covariates considered for use in DOMs addressed aspects of habitat and land cover [@fig-CovariateSummary]. 35% of studies incorporated covariates for site geometry and connectivity, such as habitat patch size or distance to other sites. Often these were included as simple covariates on colonisation or extinction, such as the distance to other sites or landscape habitat connectivity metrics [@duggan2011]. Alternatively, more complex parameterisations explicitly modelled colonisation or extinction as a spatial process dependent on patch size or the distance to occupied sites [@risk2011; @broms2016]. Several studies also included biotic interactions with other species as covariates, often where the target taxon was threatened by invasive species. These covariates effectively incorporate species interactions in DOMs without requiring the use of more complex explicitly multi-species models.
